Empirical Analysis of Performance Bottlenecks in Graph Neural Network Training and Inference with GPUs

### 2020-12-26 NEUROC/sec5-insights.tex
更新后几章

Through the extensive experiments, we propose the following key findings and suggestions for how to optimize the performance of the GNN training/inference.

@@

    \item \emph{The computational cost of a GNN layer is mainly affected by the dimensions of the input and the output hidden feature vectors.}
          Theoretically and empirically, the training/inference time and the memory usage of a GNN layer both increase \emph{linearly} with the dimensions of the input/output hidden feature vectors separately.
          GNNs are friendly to high-dimensional scenarios. 
          Algorithm engineers can use high-dimensional feature vectors to improve the expressive power of a GNN without worrying exponential growth in the training/inference time and memory usage
    \item \emph{Performance optimizations should focus on improving the efficiency of the edge calculation stage.}
          The edge calculation stage is the most time-consuming stage in most GNNs.
          \begin{itemize}
              \item If the complexity of the messaging function $\phi$ is high, the implementation of $\phi$ is critical to performance.
                    Improving its efficiency can significantly reduce the training time.
                    For example, the attention mechanism in GNNs (like GAT and GaAN) requires an extra sub-layer to calculate the attention weight of each edge.
                    Implementing the attention mechanism with specially optimized basic operators on the GPU side is a potential optimization.
              \item If the complexity of $\phi$ is low, the efficiency of the collection step and the aggregation step becomes critical.
                    The existing GNN libraries \cite{DGL, PyG, ma2019_neugraph} already introduce the \emph{fused} operator to improve their efficiency.
                    When the messaging function $\phi$ is an assignment or a scalar multiplication of the hidden feature vector of the source vertex, the libraries replace the collection, messaging, and aggregation steps with a single fused operator.
                    The fused operator calculates the aggregated vectors directly from the input hidden feature vectors, minimizing the memory footprints and overlapping the memory accessing with computation.
                    In this way, it significantly reduces the training/inference time of GNNs with low edge calculation complexity (like GCN) \cite{yan2020_characterizing_gcn, zhang2020_analysis_neugraph}.
                    However, the applicable condition of the fused operator is very restricted.
                    It does not work for $\phi$ with more complex operations like matrix multiplication.
                    A potential optimization is to develop composite CUDA kernels that can read the input hidden feature vectors and aggregate message vectors on the fly, without materializing the parameter vectors and the message vectors.          \end{itemize}
    \item \emph{The high memory usage caused by the intermediate results of the edge calculation stage limits the data scalability of the GNN training/inference.}
          The memory expansion ratios of the typical GNNs are very high, making GPUs unable to handle big graphs.
          One solution is to distribute the dataset among several GPUs and frequently swap parts of the dataset between GPUs and the main memory \cite{ma2019_neugraph}.
          Another possible solution \cite{chen2016_training_deep} comes from the deep neural network training. % checkpoints only work for GNN training.
          It only checkpoints key intermediate results during the forward propagation and re-calculates the missing results on demand during the backpropagation.
          Implementing the checkpoint mechanism is another potential optimization for GNN training.

    \item \emph{Sampling techniques can significantly reduce the training/inference time and memory usage, but its implementation is still inefficient}.
          The sampling techniques are effective under small batch sizes.
          Its current implementation brings considerable overheads when the batch size becomes large.
          Improving the efficiency of the sampling is a potential optimization.
          The sampled subgraphs are usually small.
          They cannot make full use of the computing power of a GPU.
          How to improve the GPU utilization under small batch sizes is another problem to solve.
          One possible solution is to train multiple batches asynchronously on the same GPU and use the asynchronous stochastic gradient descent to speed up the converge.

\end{enumerate}

>>>>

Through the extensive experiments, we propose the following key findings and suggestions for how to optimize the performance for GNN training and inference.

@@

    \item \emph{The computational cost of a GNN layer is mainly affected by the dimensions of the input and output hidden vectors.}
          Theoretically and empirically, the training and inference time and the memory usage of a GNN layer both increase \emph{linearly} with the dimensions of the input/output hidden vectors separately.
          GNNs are friendly to high-dimensional scenarios. 
          Algorithm engineers can use high-dimensional feature vectors to improve the expressive power of a GNN without worrying exponential growth in the training/inference time and memory usage
    \item \emph{Performance optimizations should focus on improving the efficiency of the edge calculation stage.}
          The edge calculation stage is the most time-consuming stage in most GNNs.
          \begin{itemize}
              \item If the complexity of the messaging function $\phi$ is high, the implementation of $\phi$ is critical to performance.
                    Improving its efficiency can significantly reduce training/inference time.
                    For example, the attention mechanism in GNNs (like GAT and GaAN) requires an extra sub-layer to calculate the attention weight of each edge.
                    Implementing the attention mechanism with specially optimized basic operators on the GPU side is a potential optimization direction.
              \item If the complexity of $\phi$ is low, the efficiency of the collection step and the aggregation step becomes critical.
                    The existing GNN libraries \cite{DGL, PyG, ma2019_neugraph} already introduce the \emph{fused} operator to improve their efficiency.
                    When the messaging function $\phi$ is an assignment or a scalar multiplication of the input hidden vector of the source vertex, the libraries replace the collection, messaging, and aggregation steps with a single fused operator.
                    The fused operator calculates the aggregated vectors directly from the input hidden vectors, minimizing the memory footprints and overlapping the memory accessing with computation.
                    In this way, it significantly reduces the training/inference time of GNNs with low edge calculation complexity (like GCN) \cite{yan2020_characterizing_gcn, zhang2020_analysis_neugraph}.
                    However, the applicable condition of the fused operator is very restricted.
                    It does not work for $\phi$ with more complex operations like matrix multiplication.
                    A potential optimization is to develop composite CUDA kernels that can read the input hidden vectors and aggregate message vectors on the fly, without materializing the parameter vectors and the message vectors.
         \end{itemize}
    \item \emph{The high memory usage caused by the intermediate results of the edge calculation stage limits the data scalability of GNN training/inference.}
          The memory expansion ratios of the typical GNNs are very high, making GPUs unable to handle big graphs.
          One solution is to distribute the dataset among several GPUs and frequently swap parts of the dataset between GPUs and the main memory \cite{ma2019_neugraph}.
          Another possible solution \cite{chen2016_training_deep} comes from the deep neural network training. % checkpoints only work for GNN training.
          It only checkpoints key intermediate results during the forward propagation and re-calculates the missing results on demand during the backpropagation.
          Implementing the checkpoint mechanism is another potential optimization for GNN training.

    \item \emph{Sampling techniques can significantly reduce the training/inference time and memory usage, but its implementation is still inefficient}.
          The sampling techniques are effective under small batch sizes for both training and inference.
          However, its current implementation brings considerable overheads when the batch size is not big.
          The sampled subgraphs that are usually small cannot make full use of the computing power of a GPU either.
          Improving the efficiency of the sampling techniques is a potential optimization direction.
          How to improve the GPU utilization under small batch sizes is another problem to solve.
          One possible solution is to train multiple batches asynchronously on the same GPU and use the asynchronous stochastic gradient descent to speed up the converge.
\end{enumerate}

### 2020-12-26 NEUROC/sec7-conclusion.tex
更新后几章

In this work, we systematically explore the performance bottleneck in graph neural network training and inference. 
We model the existing GNNs with the message-passing framework. 
We classify the GNNs according to their edge and vertex calculation complexities to select four typical GNNs for evaluation. 
The experimental results validate our complexity analysis. Fixing other hyper-parameters, the training time and the memory usage increase linearly with each hyper-parameter of the four GNNs. 
To find out the performance bottleneck in the training/inference time, we decompose the training/inference time per epoch on different levels. 
The time breakdown analysis indicates that the edge calculation stage and its related basic operators are the performance bottleneck for most GNNs. 
Moreover, the intermediate results produced by the edge calculation stage cause high memory usage, limiting the data scalability. 
Adopting sampling techniques can reduce the training time and the memory usage significantly. 
However, the current implementation of the sampling techniques in PyG brings considerable sampling overheads. 
The small sampled subgraphs cannot make full use of the computing power of a GPU card either. 
Our analysis indicates that the edge calculation stage should be the main target of optimizations. 
Reducing its memory usage and improving its efficiency can significantly improve the performance of the GNN training and inference. 
Based on the analysis, we propose several potential optimizations for the GNN frameworks. 
We believe that our analysis can help developers to have a better understanding of the characteristics of GNN training and inference.

>>>>

In this work, we systematically explore the performance bottlenecks in graph neural network training and inference. 
We model the existing GNNs with the message-passing framework. 
We classify the GNNs according to their edge and vertex calculation complexity to select four typical GNNs for evaluation. 
The experimental results validate our complexity analysis. Fixing other hyper-parameters, the training time and the memory usage increase linearly with each hyper-parameter of the four GNNs. 
To find out the performance bottleneck in the training/inference time, we decompose the training/inference time per epoch on different levels. 
The time breakdown analysis indicates that the edge calculation stage and its related basic operators are the performance bottleneck for most GNNs. 
Moreover, the intermediate results produced by the edge calculation stage cause high memory usage, limiting the data scalability. 
Adopting sampling techniques can reduce the training time and the memory usage significantly. 
However, the current implementation of the sampling techniques in PyG brings considerable sampling overheads. 
The small sampled subgraphs cannot make full use of the computing power of a GPU card either. 
Our analysis indicates that the edge calculation stage should be the main target of optimizations. 
Reducing its memory usage and improving its efficiency can significantly improve the performance of the GNN training and inference. 
Based on the analysis, we propose several potential optimizations for the GNN frameworks. 
We believe that our analysis can help developers to have a better understanding of the characteristics of GNN training and inference.
